---
title: "ggplot2: Handout A, The Grammar of Graphics"
author: "Eric N. Moyer"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: pdflatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: pdflatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(ggplot2)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

This is the first of a set of "Tufte Handouts" documenting my exploration of 
Hadley Wickham's book *`ggplot2`: Elegant Graphics for Data Analysis, Second 
Edition* (2016). The purpose is to record helpful bits of understanding while 
exercising my skills.

Everywhere in these handouts, the word "plot" means a plot produced by `ggplot2`.

# The Grammar of Graphics

`ggplot2` is built on a "layered grammar of graphics", which makes every plot
the product of a uniform, deeply data-driven, theory of graphics, instead of just
being another "special case" (as with most plotting systems.)

From the book: "In brief, the grammar tells us that a statistical graphic is 
a mapping from data to aesthetic attributes (colour, shape, size) of geometric 
objects (points, lines, bars)."

# Key Components of Every Plot

Every plot begins with a call to `ggplot()`. Although you would never 
do so in practice because all it gives you is an empty background, you could 
just leave it at that, and the call would complete without errors:

```{r fig-ggplot-alone}
ggplot()
```

The *first* thing every plot needs is some data, which is always in 
the form of at least one dataframe. The dataframe(s) can be provided 
in various ways; often the primary one is passed to `ggplot()`:

```{r fig-ggplot-dataonly}
ggplot(mpg)
```

However, providing only the data (and nothing else) doesn't change anything, 
because `ggplot2` doesn't know what to do with it, so the *second* thing every 
plot needs is some "aesthetic mappings", which connect variables in the data 
with "visual properties":

```{r fig-ggplot-aes}
ggplot(mpg, aes(x = displ, y = hwy))
```

We have told `ggplot2` to associate the visual property of horizontal 
location (`x`) with the variable `displ`, and vertical location with `hwy`, 
so `ggplot2` can look at the range of values for these variables and use 
implicit "scale" components to create "guides" for reading the plot, which 
in this case are tickmarks on the axes.

But nothing appears on the plot because we have not specified what kind of 
geometric object(s) to use, so the *third* thing every plot needs is a "layer" 
for rendering each observation in the data, and this is usually specified with 
a `geom` fuction:

```{r fig-ggplot-done}
ggplot(mpg, aes(displ, hwy)) + geom_point()
```

And now we have a complete plot, albeit a very simple one. `ggplot2` rendered 
each observation in the `mpg` dataframe by using the `displ` and `hwy` variables 
as the x- and y-coordinates of a point.

There is much, much more to understand, but this is the foundation.

Before leaving this first example, we note that for any geometric object, each 
of the visual properties it supports can be either mapped to a variable in the 
data or set to a constant.

used here 
(points), in addition to x and y position, also has the visual properties of 
color, shape, and size.

visual properties that 
can be used to represent a variable often include such things as size, 
shape, and color:

```{r fig-ggplot-color}
ggplot(mpg, aes(displ, hwy, color = class)) + geom_point()
```



