---
title: "ggplot2: Handout A, The Grammar of Graphics"
author: "Eric N. Moyer"
date: "`r Sys.Date()`"
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: pdflatex
  tufte::tufte_html: default
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: pdflatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(ggplot2)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

This is the first of a set of "Tufte Handouts" documenting my exploration of 
Hadley Wickham's book *`ggplot2`: Elegant Graphics for Data Analysis, Second 
Edition* (2016). The purpose is to record helpful bits of understanding while 
exercising my skills.

Everywhere in these handouts, the word "plot" means a plot produced by `ggplot2`.

# The Grammar of Graphics

`ggplot2` is built on a "layered grammar of graphics", which makes every plot
the product of a uniform, deeply data-driven, theory of graphics, instead of 
a "special case".

From the book: "In brief, the grammar tells us that a statistical graphic is 
a mapping from data to aesthetic attributes (colour, shape, size) of geometric 
objects (points, lines, bars)."

# Three Things Every Plot Needs

Every plot begins with a call to `ggplot()`. You *could* just leave it at that, 
and the call would complete without errors, but all you would get is an empty 
background:

```{r fig-ggplot-alone}
ggplot()
```

So the *first* thing every plot needs is data, which is always in 
the form of at least one dataframe. The dataframe(s) can be provided 
in various ways; often the primary one is passed to `ggplot()`:

```{r fig-ggplot-dataonly}
ggplot(mpg)
```

This still doesn't get us very far, because we haven't told 
`ggplot2` what to do with the data, so the *second* thing every 
plot needs is "aesthetic mappings" to connect variables in the data 
with "visual properties":

```{r fig-ggplot-aes}
ggplot(mpg, aes(x = displ, y = hwy))
```

Now we have told `ggplot2` to associate the visual property of horizontal
location (`x`) with the variable `displ`, and vertical location with `hwy`, so
it can look at the ranges of these variables and use implicit "scale" components
to create "guides" for reading the plot, which in this case are tickmarks on the
axes.

However, nothing appears on the plot yet because we have not specified what kind of 
geometric object(s) to use, so the *third* thing every plot needs is at least one "layer" 
that renders the observations in the data, and this is usually specified with 
a `geom` fuction:

```{r fig-ggplot-done}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```

Now we have a complete plot. `ggplot2` rendered each observation in the `mpg`
dataframe by using the `displ` and `hwy` variables as the x- and y-coordinates
of a point.

There is much, much more to understand, but this is the foundation.

Note that for any geometric object, each of the visual properties it supports
can be either mapped to a variable in the data or set to a constant. Any
properties you don't use are set to default constants by `ggplot2`. The point
geom used above supports not only x and y, but also color, shape, and size.
Since we did not specify otherwise, `ggplot2` showed the points as 3mm black
circles.

# The Six Components of the Grammar

Ffffff
